Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-06 paper/paper.tex
correctness of liveness analysis
Fixed a minor in liveness GC for closures

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\blankout}[1]{}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}

\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma} 
\begin{proof}
Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
grammar $G$  such that  the machine will  halt on  $w$ if and  only if
$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
unrestricted productions  in Definition~\ref{def:specialgrammar}.

@@

Assume that  the TM  starts in a  state $S_\mathit{init}$ with  a tape
$cw$  and the  head   positioned on  the symbol  $c$. Then  the the
sentential  form   corresponding  to  the   initial  configuration  is
$\nont{\mathit{init}}{c}w$ (we can assume  that there is a production
$\nont{}{}  \rightarrow \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of  the grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

@@

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.  
\end{proof}

We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.

>>>>

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\cblue}[1]{{\color{blue}{#1}}}
\newcommand{\blankout}[1]{}

@@

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}
\cblue{
\begin{lemma}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The  decision problem
$\alpha \in \mathscr{L}(G)$ is undecidable.
\end{lemma} 
\begin{proof}
Given a  Turing machine  and an input  $w\in (1+0)^*$, we  construct a
grammar $G$  such that  the machine will  halt on  $w$ if and  only if
$\epsilon \in  \mathscr{L}(G)$. The grammar includes the  fixed set of
unrestricted productions  in Definition~\ref{def:specialgrammar}.

@@

Assume that  the TM starts  in a  state $S_\mathit{init}$ with  a tape
$cw$ and  the head positioned on  the symbol $c$. Then  the sentential
form    corresponding     to    the    initial     configuration    is
$\nont{\mathit{init}}{c}w$ (we  can assume that there  is a production
$\nont{}{} \rightarrow  \nont{\mathit{init}}{c}w$, where  \nont{}{} is
the start symbol of the  grammar). Further correspondences between the
Turing machine moves and the grammar productions are as follows:

@@

After  the Turing machine  has halted,  there are  further ``cleanup''
derivations  that derive  $\epsilon$ only  if the  invariant  has been
maintained  so far.  For every  symbol  $c$, we introduce  a  non-terminal
$\nont{\mathit{final}}{c}$ where $\nont{\mathit{final}}{}$ is the
final state of the Turing machine. We add
productions $\nont{\mathit{final}}{c} \rightarrow \acar
\nont{\mathit{final}}{c}$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}$ for cleaning up the \bcar\ and \bcdr\ symbols on
the left of the head and $\nont{\mathit{final}}{c} \rightarrow 
\nont{\mathit{final}}{c}\bcar$ and $\nont{\mathit{final}}{c} \rightarrow \acdr
\nont{\mathit{final}}{c}\bcdr$ for cleaning up \acar\ and \acdr\ on the
right of the tape head. This completes the reduction.  
\end{proof}
}%end of cblue
We circumvent the problem  of undecidability by over- approximating the
CFG by non-deterministic finite state automata (NFA). The NFA are then
simplified using  the $\hookrightarrow$ rules.  Finally the simplified
NFAs are converted to DFAs.
